Relational programming in Smalltalk

### 2019-03-06 booklet/article.tex
Description of `Succeed` and `Fail` goal for the booklet.

@@

\section{Introduction}

@@

\section{Resolution methods}

@@

\begin{description}

\item[by refutation] Let $\alpha$ be a sentence in \emph{conjunctive normal
form} (CNF for short) and $M(\alpha)$ the set of models that satisfy $\alpha$
-- recall that a \emph{model} is a set of assignments of values to free
variables in $\alpha$ that make the whole sentence true.  We say that, on one
hand, $\alpha$ is \textit{valid} if it is true in \textit{all} models; on the
other hand, $\alpha$ is \textit{satisfiable} if it is true in \textit{some}
model.

@@

\item[by unification] is a the process introduced in \citep{robinson_unif} and
improved by \citep{Martelli:1982:EUA:357162.357169}; it consists of solving
\textit{equations among symbolic expressions}. A \textit{solution} is denoted
as a \textit{substitution} $\theta$, namely a mapping that assigns a symbolic
values to free variables; let $G$ be a set of equations, unification proceeds
according to the following rules:
\begin{description}
\item[delete] $G \cup \lbrace t = t \rbrace \rightarrow G$
\item[decompose] $G \cup \lbrace f(s_{0}, \ldots, s_{k}) = f(t_{0}, \ldots, t_{k})\rbrace$ entails
$$G \cup \lbrace s_{0}=t_{0},\ldots, s_{k}=t_{k} \rbrace$$
\item[conflict] if $f\neq g \vee k\neq m$ then $$G \cup \lbrace f(s_{0}, \ldots, s_{k}) = g(t_{0}, \ldots, t_{m})\rbrace \rightarrow \,\perp$$
\item[eliminate] if $x \not\in vars(t)$ and $x \in vars(G)$ then $$G \cup \lbrace x = t\rbrace \rightarrow G\lbrace x \mapsto t\rbrace \cup \left\lbrace x \triangleq t\right\rbrace $$
\item[occur check] if $x \in vars(f(s_{0},\ldots,s_{k}))$ then $$G \cup \lbrace x = f(s_{0}, \ldots, s_{k})\rbrace \rightarrow \,\perp;$$
without it, generating a $\theta$ is a
\emph{recursive enumerable} problem.
\end{description}

\end{description}

\begin{example}
Let $x$ and $y$ be free variables, the set
$$\lbrace cons(x,cons(x,nil)) = cons(2,y)\rbrace$$
has solution $\theta = \lbrace x \mapsto 2, y \mapsto cons(2,nil) \rbrace$;
on the contrary, the set
$$ \lbrace y = cons(2,y) \rbrace $$
has no \textit{finite} solution, nonetheless
$$\theta = \lbrace y \mapsto cons(2,cons(2,cons(2,...))) \rbrace$$
is a solution upto \textit{bisimulation}
\citep{10.1007/BFb0017309, DBLP:books/daglib/0067019}.
\end{example}

\section{Conclusions}

A quine generator is described in \citep{Byrd:2012:MLU:2661103.2661105}.

>>>>

\newcommand{\stcode}[1]{\begin{minted}[fontsize=\footnotesize]{smalltalk} #1 \end{minted}}

@@

\section*{Introduction}

@@

\tableofcontents

\section{Bootstrapping $\mu$Kanren}

We believe that an unusual development process can be done in Smalltalk and it
is the one that had driven and currently drives the present description.
Instead of creating classes first, we start by looking at some \textit{existing
object} that can be a good candidate to take into account our messages; in
sight of this approach, we give the fundamentals of $\mu$Kanren -- more
generally, for a class of logic programming languages -- starting from the two
nice boolean objects.

\subsection{\texttt{Succeed} and \texttt{Fail} classes}

Let us consider the object \Verb|false| whose responsibility is to represent
\textit{falsehood}, hence it should be a good candidate to provide us something
related to this from the \textit{relational} point of view,
\begin{minted}[fontsize=\footnotesize]{smalltalk}
false asGoal
\end{minted}
namely, \Verb|false| can build for us an object that represents a relation that
cannot be satisfied at all responding to the message selector \Verb|asGoal| --
don't be scared by multiple names, we freely interchange the terms
\textit{goal} and \textit{relation} postponing to later sections the
description of their technical difference. Moreover, let us 
assume that the infrastructure that actually build results is given and implemented
by the \textit{oracle block} \texttt{B := [:g | g solutions all]}, we write our first test case:
\begin{minted}[fontsize=\footnotesize]{smalltalk}
GoalTest>>#testFail
    self 
        assert: false asGoal solutions all 
        equals: {}
\end{minted}
Now we get within the \Verb|False| class to add the implementation of the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
False>>#asGoal
    ^ Goal fail
\end{minted}
and of the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Goal>>#fail
    ^ Fail new
\end{minted}
such that to say that \verb|Fail| denotes the impossibility to satisfy a relation
is the same to code the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Fail>>#onState: aState 
    ^ Chain bottom
\end{minted}
which will be used within the message \Verb|solutions| sent in the oracle block
\Verb|B| -- a lot remains to be explained, such as the \Verb|Chain| class and
streams in particular, all in what follows.

In parallel, we reproduce for the object \Verb|true| the steps did for
the object \Verb|false|; first,
\begin{minted}[fontsize=\footnotesize]{smalltalk}
GoalTest>>#testSucceed
    self 
        assert: true asGoal solutions all 
        equals: {Var tautology}
\end{minted}
second,
\begin{minted}[fontsize=\footnotesize]{smalltalk}
True>>#asGoal
    ^ Goal succeed
\end{minted}
third,
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Goal>>#succeed
    ^ Succeed new
\end{minted}
fourth,
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Succeed>>#onState: aState 
    ^ Chain with: aState
\end{minted}

\section{Theoretical backgrounds}

@@

\subsection{Resolution by refutation}

Let $\alpha$ be a sentence in \emph{conjunctive normal
form} (CNF for short) and $M(\alpha)$ the set of models that satisfy $\alpha$
-- recall that a \emph{model} is a set of assignments of values to free
variables in $\alpha$ that make the whole sentence true.  We say that, on one
hand, $\alpha$ is \textit{valid} if it is true in \textit{all} models; on the
other hand, $\alpha$ is \textit{satisfiable} if it is true in \textit{some}
model.

@@

\subsection{Resolution by unification}

This method was introduced in \citep{robinson_unif} and
improved by \citep{Martelli:1982:EUA:357162.357169}; it consists of solving
\textit{equations among symbolic expressions}. A \textit{solution} is denoted
as a \textit{substitution} $\theta$, namely a mapping that assigns a symbolic
values to free variables; let $G$ be a set of equations, unification proceeds
according to the following rules:
\begin{description}
\item[delete] $G \cup \lbrace t = t \rbrace \rightarrow G$
\item[decompose] $G \cup \lbrace f(s_{0}, \ldots, s_{k}) = f(t_{0}, \ldots, t_{k})\rbrace$ entails
$$G \cup \lbrace s_{0}=t_{0},\ldots, s_{k}=t_{k} \rbrace$$
\item[conflict] if $f\neq g \vee k\neq m$ then $$G \cup \lbrace f(s_{0}, \ldots, s_{k}) = g(t_{0}, \ldots, t_{m})\rbrace \rightarrow \,\perp$$
\item[eliminate] if $x \not\in vars(t)$ and $x \in vars(G)$ then $$G \cup \lbrace x = t\rbrace \rightarrow G\lbrace x \mapsto t\rbrace \cup \left\lbrace x \triangleq t\right\rbrace $$
\item[occur check] if $x \in vars(f(s_{0},\ldots,s_{k}))$ then $$G \cup \lbrace x = f(s_{0}, \ldots, s_{k})\rbrace \rightarrow \,\perp;$$
without it, generating a $\theta$ is a
\emph{recursive enumerable} problem.
\end{description}

\begin{example}
Let $x$ and $y$ be free variables, the set
$$\lbrace cons(x,cons(x,nil)) = cons(2,y)\rbrace$$
has solution $\theta = \lbrace x \mapsto 2, y \mapsto cons(2,nil) \rbrace$;
on the contrary, the set
$$ \lbrace y = cons(2,y) \rbrace $$
has no \textit{finite} solution, nonetheless
$$\theta = \lbrace y \mapsto cons(2,cons(2,cons(2,...))) \rbrace$$
is a solution upto \textit{bisimulation}
\citep{10.1007/BFb0017309, DBLP:books/daglib/0067019}.
\end{example}
\section{Conclusions}

A quine generator is described in \citep{Byrd:2012:MLU:2661103.2661105} and
we aim to reproduce it.
